working title Compatibility system and stygma size of pollen recipient as main predictors of heterospecific pollen effect

### 2018-12-07 Manuscript_draft/Manuscript_draft.Rmd
paragraph 3

Recent studies have advanced in the ecological understanding of heterospecific pollen effect @morales2008 @ashman2013 @arceo2016. A general overview of foreign pollen arrival is that it can play an important role on species fitness but seems to be context dependent and not always produce a decrease in fitness @morales2008. Part of this unpredictability is due to the enormous variability of foreing pollen transferred in nature, where levels between 0 and 75 percent are seen, but most commonly values ranges between 0 and 20 percent of the total pollen load, being the generalist species the ones that receive greater loads of heterospecific pollen @bartomeus2008 @montgomery2012 @ashman2013 @fang2013. Although heterospecific pollen quantity is fundamental to understand the outcome of the interaction so is the different traits of both pollen donor and recipient. @ashman2013 postulated the first predictive framework for traits of heterospecific pollen effect, where different traits such as compatibility system and pollen size among others seems to be crucial to understand foreing pollen effect. Moreover, in @tong2016 an assymetric effect was shown in a crossing experiment between 6 species of the genus *Pedicularis* where the pollen of long styled species was able to grow the full length of the style on short styled species but not viceversa. Despite these recent caveats, we still lack empirical evidence to affirm what are the main traits that drive heterospecific pollen effect for both pollen donor and recipient at seed production level. Interestingly, to comprehend how these traits interact is also crucial to look at the phylogenetic relatedness of the species. There is a considerable amount of literature of crosses between close related species @brown2001 @arceo2016can @tong2016 but few works focused on heterospecific pollen of far related species which is a more realistic feature in natural systems due to a less likely niche overlap (REF). Although close related species seems to affect greatly species fitness @arceo2016 the importance of heterospecific pollen effect of far and close related species remains to be explored at community level.

@@

Traditionally,

. For this reason, controlled pollination experiments where hand pollination is applied instead of using pollinators as pollen vectors have simplified this task. However, other issues may arise such as overestimation of the effect (REFs), different methods across studies or lack of descriptive methodology for reproducible work (REFs). One of the most common ways to estimate heterospecific pollen effect is through hand pollination with mixes of 50% conspecific pollen and 50% heterospecific pollen (REF).

I would like to add that the experiments focus on two proxies of effect prezygotic and postzygotic. Why focus on postzygotic? Is the final stage where we can see the effect. Further studies should also study germination rates.

Traditionally heterospecific pollen effect has focused its attention on different pollen donors as a main driver of different effect. However in this article we want to emphasize that this is true for the cases that the species are higly close related where pollen recognition can take place (eg hybridization) but not when this pollen is from less closely related species which the main driver of effect is determined by the reproductive biology of the female part of the plant(compatibility system, stigma type, stigma area and number of ovules).

@@

The great environmental variability in natural systems and complexity of floral structures make heterospecific pollination studies a daunting task. Although plant-pollinator network and pollen network studies can give a first picture of the importance of foreign pollen is necessary to address how its effect is shaped with both traits and relatedness of the species. For this reason, in this study we have created an artificial co-flowering community with 10 species belonging to three different families where we try to test the following questions: 1) Does heterospecific pollen reduce seed set, if so, 2) Does heterospecific pollen effect depend on the relatedness of the species, 3) Does heterospecific pollen effect depend on any floral trait?

>>>>

Recent studies have advanced in the ecological understanding of heterospecific pollen effect @morales2008 @ashman2013 @arceo2016. A general overview of foreign pollen arrival is that it can play an important role on species fitness but seems to be context dependent and not always produce a decrease in fitness @morales2008. Part of this unpredictability is due to the enormous variability of foreing pollen transferred in nature, where levels between 0 and 75 percent are seen, but most commonly values ranges between 0 and 20 percent of the total pollen load, being the generalist species the ones that receive greater loads of heterospecific pollen @bartomeus2008 @montgomery2012 @ashman2013 @fang2013. Although heterospecific pollen quantity is fundamental to understand the outcome of the interaction so is the different traits of both pollen donor and recipient. @ashman2013 postulated the first predictive framework for traits of heterospecific pollen effect, where different traits such as compatibility system and pollen size among others seems to be crucial to understand foreing pollen effect. Moreover, in @tong2016 an assymetric effect was shown in a crossing experiment between 6 species of the genus *Pedicularis* where the pollen of long styled species was able to grow the full length of the style on short styled species but not viceversa. Despite these recent caveats, we still lack empirical evidence to affirm what are the main traits that drive heterospecific pollen effect for both pollen donor and recipient at seed production level. Interestingly, to comprehend how these traits interact is also crucial to look at the phylogenetic relatedness of the species. There is a considerable amount of literature of crosses between close related species @brown2001 @arceo2016can @tong2016 but few works focused on heterospecific pollen of far related species. Although the effect of close related species is predicted to be greater @ashman2013 the presence of pollen of non related species on multiple species @arceo2016 and the higher chances to coexist with a species that has less niche overlap (Ref) make foreign pollen from far related species also an important subject of study in order to understand the importance of heterospecific pollen in natural systems. Notwithstanding, the effect of heterospecific pollen of far and close related species at community level remains to be explored.

@@

Interestingly, incompatibility system seems to play an important role in foreign pollen effect where species that are self incompatible would have stronger barriers towards heterospecific pollen than self compatible species.  The type of incompatibility, sporophytic or gametophytic is related with the place of pollen recognition where the former take place at the sitgma level and the latter occurs within the style, this last late acting pollen recognition mechanism is associated with greater negative effect (REF). Remarkably, there is a great variability in mating systems across populations @whitehead2018 and therefore predict an effect of foreign pollen is a bit obscured by the variability within species, however species that are strong selfers or strong outcrossers have less variablity in mating systems and predictions of effect could be more realistic (see figure 1 from @whitehead2018). Moreover, other traits such as number of pollen grains per flower and number of ovules have been tradittionally associated with the type of incompatibility system where species with higher pollen ovule ratios are predicted to be xenogamous and species with low pollen ovule ratios autogamous (REFS).

@@

The great environmental variability in natural systems and complexity of floral structures make heterospecific pollination studies a daunting task. Moreover, variation in sampling effort have been shown to be determinant to characterize pollen transfer interactions @arceo2018. Although plant-pollinator network and pollen network studies can give a first picture of the importance of foreign pollen is necessary to address how its effect is shaped with both traits and relatedness of the species. For this reason, in this study we have created an artificial co-flowering community with 10 species belonging to three different families where we try to test the following questions: 1) Does heterospecific pollen reduce seed set, if so, 2) Does heterospecific pollen effect depend on the relatedness of the species, 3) Does heterospecific pollen effect depend on any floral trait?

### 2018-12-07 Manuscript_draft/Manuscript_draft.tex
paragraph 3

Recent studies have advanced in the ecological understanding of
heterospecific pollen effect Morales and Traveset (2008) Ashman and
Arceo-Gómez (2013) Arceo-Gómez and Ashman (2016). A general overview of
foreign pollen arrival is that it can play an important role on species
fitness but seems to be context dependent and not always produce a
decrease in fitness Morales and Traveset (2008). Part of this
unpredictability is due to the enormous variability of foreing pollen
transferred in nature, where levels between 0 and 75 percent are seen,
but most commonly values ranges between 0 and 20 percent of the total
pollen load, being the generalist species the ones that receive greater
loads of heterospecific pollen Bartomeus et al. (2008) Montgomery and
Rathcke (2012) Ashman and Arceo-Gómez (2013) Fang and Huang (2013).
Although heterospecific pollen quantity is fundamental to understand the
outcome of the interaction so is the different traits of both pollen
donor and recipient. Ashman and Arceo-Gómez (2013) postulated the first
predictive framework for traits of heterospecific pollen effect, where
different traits such as compatibility system and pollen size among
others seems to be crucial to understand foreing pollen effect.
Moreover, in Tong and Huang (2016) an assymetric effect was shown in a
crossing experiment between 6 species of the genus \emph{Pedicularis}
where the pollen of long styled species was able to grow the full length
of the style on short styled species but not viceversa. Despite these
recent caveats, we still lack empirical evidence to affirm what are the
main traits that drive heterospecific pollen effect for both pollen
donor and recipient at seed production level. Interestingly, to
comprehend how these traits interact is also crucial to look at the
phylogenetic relatedness of the species. There is a considerable amount
of literature of crosses between close related species Brown and
Mitchell (2001) Arceo-Gómez et al. (2016) Tong and Huang (2016) but few
works focused on heterospecific pollen of far related species which is a
more realistic feature in natural systems due to a less likely niche
overlap (REF). Although close related species seems to affect greatly
species fitness Arceo-Gómez and Ashman (2016) the importance of
heterospecific pollen effect of far and close related species remains to
be explored at community level.

@@

Traditionally,

. For this reason, controlled pollination experiments where hand
pollination is applied instead of using pollinators as pollen vectors
have simplified this task. However, other issues may arise such as
overestimation of the effect (REFs), different methods across studies or
lack of descriptive methodology for reproducible work (REFs). One of the
most common ways to estimate heterospecific pollen effect is through
hand pollination with mixes of 50\% conspecific pollen and 50\%
heterospecific pollen (REF).

I would like to add that the experiments focus on two proxies of effect
prezygotic and postzygotic. Why focus on postzygotic? Is the final stage
where we can see the effect. Further studies should also study
germination rates.

Traditionally heterospecific pollen effect has focused its attention on
different pollen donors as a main driver of different effect. However in
this article we want to emphasize that this is true for the cases that
the species are higly close related where pollen recognition can take
place (eg hybridization) but not when this pollen is from less closely
related species which the main driver of effect is determined by the
reproductive biology of the female part of the plant(compatibility
system, stigma type, stigma area and number of ovules).

@@

The great environmental variability in natural systems and complexity of
floral structures make heterospecific pollination studies a daunting
task. Although plant-pollinator network and pollen network studies can
give a first picture of the importance of foreign pollen is necessary to
address how its effect is shaped with both traits and relatedness of the
species. For this reason, in this study we have created an artificial
co-flowering community with 10 species belonging to three different
families where we try to test the following questions: 1) Does
heterospecific pollen reduce seed set, if so, 2) Does heterospecific
pollen effect depend on the relatedness of the species, 3) Does
heterospecific pollen effect depend on any floral trait?

@@

@@

>>>>

Recent studies have advanced in the ecological understanding of
heterospecific pollen effect Morales and Traveset (2008) Ashman and
Arceo-Gómez (2013) Arceo-Gómez and Ashman (2016). A general overview of
foreign pollen arrival is that it can play an important role on species
fitness but seems to be context dependent and not always produce a
decrease in fitness Morales and Traveset (2008). Part of this
unpredictability is due to the enormous variability of foreing pollen
transferred in nature, where levels between 0 and 75 percent are seen,
but most commonly values ranges between 0 and 20 percent of the total
pollen load, being the generalist species the ones that receive greater
loads of heterospecific pollen Bartomeus et al. (2008) Montgomery and
Rathcke (2012) Ashman and Arceo-Gómez (2013) Fang and Huang (2013).
Although heterospecific pollen quantity is fundamental to understand the
outcome of the interaction so is the different traits of both pollen
donor and recipient. Ashman and Arceo-Gómez (2013) postulated the first
predictive framework for traits of heterospecific pollen effect, where
different traits such as compatibility system and pollen size among
others seems to be crucial to understand foreing pollen effect.
Moreover, in Tong and Huang (2016) an assymetric effect was shown in a
crossing experiment between 6 species of the genus \emph{Pedicularis}
where the pollen of long styled species was able to grow the full length
of the style on short styled species but not viceversa. Despite these
recent caveats, we still lack empirical evidence to affirm what are the
main traits that drive heterospecific pollen effect for both pollen
donor and recipient at seed production level. Interestingly, to
comprehend how these traits interact is also crucial to look at the
phylogenetic relatedness of the species. There is a considerable amount
of literature of crosses between close related species Brown and
Mitchell (2001) Arceo-Gómez et al. (2016) Tong and Huang (2016) but few
works focused on heterospecific pollen of far related species. Although
the effect of close related species is predicted to be greater Ashman
and Arceo-Gómez (2013) the presence of pollen of non related species on
multiple species Arceo-Gómez and Ashman (2016) and the higher chances to
coexist with a species that has less niche overlap (Ref) make foreign
pollen from far related species also an important subject of study in
order to understand the importance of heterospecific pollen in natural
systems. Notwithstanding, the effect of heterospecific pollen of far and
close related species at community level remains to be explored.

@@

Interestingly, incompatibility system seems to play an important role in
foreign pollen effect where species that are self incompatible would
have stronger barriers towards heterospecific pollen than self
compatible species. The type of incompatibility, sporophytic or
gametophytic is related with the place of pollen recognition where the
former take place at the sitgma level and the latter occurs within the
style, this last late acting pollen recognition mechanism is associated
with greater negative effect (REF). Remarkably, there is a great
variability in mating systems across populations Whitehead et al. (2018)
and therefore predict an effect of foreign pollen is a bit obscured by
the variability within species, however species that are strong selfers
or strong outcrossers have less variablity in mating systems and
predictions of effect could be more realistic (see figure 1 from
Whitehead et al. (2018)).

@@

The great environmental variability in natural systems and complexity of
floral structures make heterospecific pollination studies a daunting
task. Moreover, variation in sampling effort have been shown to be
determinant to characterize pollen transfer interactions Arceo-Gómez et
al. (2018). Although plant-pollinator network and pollen network studies
can give a first picture of the importance of foreign pollen is
necessary to address how its effect is shaped with both traits and
relatedness of the species. For this reason, in this study we have
created an artificial co-flowering community with 10 species belonging
to three different families where we try to test the following
questions: 1) Does heterospecific pollen reduce seed set, if so, 2) Does
heterospecific pollen effect depend on the relatedness of the species,
3) Does heterospecific pollen effect depend on any floral trait?

@@

\hypertarget{ref-arceo2018}{}
Arceo-Gómez, G., C. Alonso, T.-L. Ashman, and V. Parra-Tabla. 2018.
Variation in sampling effort affects the observed richness of
plant--plant interactions via heterospecific pollen transfer:
Implications for interpretation of pollen transfer networks. American
journal of botany 105:1601--1608.

@@

\hypertarget{ref-whitehead2018}{}
Whitehead, M. R., R. Lanfear, R. J. Mitchell, and J. D. Karron. 2018.
Plant mating systems often vary widely among populations. Frontiers in
Ecology and Evolution 6:38.
